---
title: "Exercise-10"
format: html
editor: visual
---

# **Exercise 10**

## **Practice ANOVA**

First I am going to load the necessary libraries we may need (identified from the previous class which we did this same exercise)

```{r}
library(tidyverse)
library(skimr)
library(broom)
library(infer)
library(emmeans)
library(sjPlot)
```

First, loading the ANOVET dataset,

```{r}

f <- "https://raw.githubusercontent.com/difiore/ada-datasets/main/AVONETdataset1.csv"
d <- read_csv(f)
```

I am first going to extract only the mentioned variables, which are : **Species1**, **Family1**, **Order1**, **Beak.Length_Culmen**, **Beak.Width**, **Beak.Depth**, **Tarsus.Length**, **Wing.Length**, **Tail.Length**, **Mass**, **Habitat**, **Migration**, **Trophic.Level**, **Trophic.Niche**, **Min.Latitude**, **Max.Latitude**, **Centroid.Latitude**, **Primary.Lifestyle**, and **Range.Size**

```{r}
d <- d %>%
  select(
    Species1, Family1, Order1,
    Beak.Length_Culmen, Beak.Width, Beak.Depth,
    Tarsus.Length, Wing.Length, Tail.Length, Mass,
    Habitat, Migration, Trophic.Level, Trophic.Niche,
    Min.Latitude, Max.Latitude, Centroid.Latitude,
    Primary.Lifestyle, Range.Size
  )
d
```

Now, I am going to do a bit of exploratory data analysis with this dataset. I will use skim() which gives quick descriptive statistics and missingness.And, glimpse() which helps identify variable classes.

```{r}
skim(d)
glimpse(d)
```

So from above, using glimpse(), i can see which of the variables are categorical and which are numeric,

Categorical variables: Species1, Family1, Order1, Habitat, Trophic.Level,Trophic.Niche,Primary.Lifestyle, Migration (stored numerically, but conceptually categorical)

Numeric variables: Beak.Length_Culmen, Beak.Width, Beak.Depth,Tarsus.Length,Wing.Length,Tail.Length, Mass, Min.Latitude, Max.Latitude, Centroid.Latitude, Range.Size

## Challenge 1: One-Factor ANOVA and Inference
